### 2014-07-27 paper.tex
updated data, text and citations

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{RNA-Seq Assembly Discovers Many Splice Variants}
}
% Insert Author names, affiliations and corresponding author email.
\\
Likit Preeyanon$^{1}$, 
Jerry B. Dodgson$^{1}$
Hans Cheng$^{2}$, 
C. Titus Brown$^{3,1 \ast}$
\\
\bf{1} Microbiology and Molecular Genetics, Michigan State University, East Lansing, MI, USA.
\\
\bf{2} Avian Disease and Oncology Laboratory, East Lansing, MI, USA.
\\
\bf{3} Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, USA.
\\
$\ast$ E-mail: ctb@msu.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Comparison of spliced reads and reference gene annotations has been
successfully used to discover alternative splicing in model
organisms. However, the method cannot be applied in organisms without
high-quality reference genome and gene annotations. In this article,
we introduce a pipeline, based on \emph{de novo} assembly, that
constructs gene models with splice variants.  The pipeline uses a
technique called local assembly that enhances the sensitivity of
alternative splicing detection.  We demonstrate that the pipeline
detects many novel splice variants in RNA-Seq data from chicken
spleen.  Many of these splice variants; however, are only detected by our
pipeline and not Cufflinks.  This indicates that the pipeline can be
used to facilitate splice variant detection from RNA-Seq.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

% @CTB: aren't some splice variants detected only by Cufflinks and not by
% gimme, too?
% @CTB: for me to check, do Tophat and cufflinks only detect canonical splice?
% @CTB: ask Olga for references to single-cell splice stuff.

@@

%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina reads),
%it is nokt feasible to assess expression isoforms and their structures without mapping reads
%to a reference annotation or reconstructing a full-length transcript from short reads.

@@

Using a similar approach, Pickrell \emph{et al} identified more than
150,000 novel canonical splice junctions in lymphoblastoid cells.  The
study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests
tissue-specific expression of isoforms\cite{Pickrell:2010gt}.  A
majority of unannotated isoforms found in both studies are from
alternative splicing of annotated exons. Only a small fraction of
isoforms contain previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman \emph{et al} used Scripture, a tool
employing similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

Although these mapping-based methods have been useful in detecting
both splice junctions and isoforms, they rely heavily on a reference
genome.  Hence, it is not necessarily practical to apply these methods
to organisms lacking a high-quality reference genome.  This limitation
can be overcome by \emph{de novo} assembly of short reads.

A number of \emph{de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a \emph{de novo} assembly
approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior than other \emph{de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses {\em de novo} assembly
to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
apply a technique we call ``local assembly'' that enhances the sensitivity
of alternative isoform detection by Oases.  The results show that the
pipeline can detect more isoforms than Oases-M and can detect isoforms
not found by Cufflinks.  We also showed that transcripts reconstructed
from \emph{de novo} assembly and mapping-based approach can be merged
to build more complete gene models.

@@

We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
assemblers to construct transcript fragments from four entire Illumina
GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
Materials).  In the assembly, we used multiple distinct k-mer values
for Velvet to sensitively recover as many different isoforms as possible
\cite{Schulz:2012je}.  We chose k-mers between 21 and 31,
and recovered between 340,000 and 370,000 unique sequences from
each data set (see Table~\ref{unique_sequences}, and Materials and
Methods).  These unique sequences represented an unknown number of
true genes, due to fragmentation from low coverage and incomplete
assembly.

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
% @CTB why are the global ones not included in the final gene models?
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (@cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

% @CTB how many reads are discarded in the process of mapping?
% @CTB what does local assembly actually do to the DBG??
% @CTB what happens if you ask how many unique regions there are not
%   represented in the final gene models?

@@

The above approaches recovered transcript fragments, but not entire genes
(figure in suppl?)
To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.
% @CTB what does this mean? ``not fully evaluated''?
We merged transcripts from our global assembly, above,
with Oases-M using a k-mer size of 27, and compared them with the
unmerged transcripts.  We then cross-validated using publicly
available ESTs, which were not used in our assembly.
The results show that
Oases-M and the unmerged assembly share about 104,413 (95\%) of the predicted
splice junctions, with 5\% disjoint.  Of these 5\%,
approximately 421 (6\%) of the Oases-M-specific splice junctions are
independently supported by ESTs, while 1,607 (19\%) of the unmerged splice
junctions are supported by ESTs.
% @CTB change from percent to total number, with percent in parantheses
This suggests that Oases-M probably
discards a number of real splice variants, although the unmerged
assembly is also missing some found by Oases-M.

% @CTB why would Oases-M create ``good'' splice variants from the unmerged assembly?
% @CTB does the math for table three work out here?
% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a general merging tool?

@@

We used an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
20,439 gene models containing 25,859 isoforms from our global
assembly; 21,963 gene models containing 25,859 from our local
assembly; and 20,922 gene models containing 27,732 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

% (more here - what? @CTB)

@@

Our pipeline predicts many gene models and isoforms after assembly.
We validated these gene models in several different ways.  XX more!

@@

We used bowtie to map single-end reads from the source datasets to the
transcripts.  More than 80\% of the original reads could be aligned to
the transcripts, demonstrating that we did not lose a significant
amount of information during the merge process
(Table~\ref{single-end_map}).
% @CTB compared to what?
More importantly, we also mapped
paired-end reads from technical replicates to the same gene models,
and found that more than 85\% of the paired-ends mapped concordantly
the gene models (Table~\ref{paired-end_map}).  Most of the reads that
did not map were either highly erroneous or contained low-complexity
artifactual sequence that probably originated from sample processing
and reverse transcription (@cite).
% @CTB do we have something we can cite for typical % mapping?  Maybe Trinity
% paper?  How about a citation that suggests that there's a lot of crap
% in mRNAseq?
Thus the merged gene models produced by Gimme represent the significant
majority of the assemblable data.

@@

Of 100,411 splice junctions from the gene models, 95 (0.09\%)
junctions have no spliced reads and only 448 (0.45\%) junctions
have fewer than 4 spliced reads (Fig.~\ref{cdf_single_splice}).
% @CTB how do you get junctions with no spliced reads!?
More than 99\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!

@@

Of the 100,411 splice junctions in our gene models, 81,838 (81.50\%)
are supported by ESTs or mRNAs from Genbank.  This is especially
surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues.  Note that
this cross-validation suggests that the 18,573 novel splice junctions
{\em not} seen in publicly available ESTs and mRNAs are also likely to
be real splice junctions from spleen.
% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?

@@

Cufflinks finds 109,641 splice junctions, and Gimme finds 100,411
splice junctions.  88,289 of them are in common.  (Venn diagram.)

Both Cufflinks and Gimme find approximately 40-50\% of the genes and
80-90\% of the splice junctions present in the ENSEMBL gene models
for chicken.  Cufflinks performs about 10\% better in both cases
than Gimme, demonstrating that Cufflinks has higher sensitivity
in recovering ENSEMBL gene models.

The ENSEMBL pipeline does not, however, include a large number of
splice junctions from ESTs (99,755) or mRNAs (13,641).  Cufflinks and
Gimme each recover about 10\% of these, with more than 2/3 of these
recovered by both Cufflinks and Gimme.  This indicates that both Gimme
and Cufflinks are equally adept at recovering novel splice junctions.

% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?

@@

When we apply Gimme and Cufflinks to a publicly available mouse
mRNAseq data set, Gimme and Cufflinks recover approximately the same
number of splice junctions already known from ENSEMBL.  However, Gimme
recovers a substantial number of additional splice variants beyond
Cufflinks and ENSEMBL both.

% @CTB is the mouse data set from the trinity paper?
% @CTB are most of these splice junctions > 3 coverage?

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity to exon-exon junctions in our gene
models by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 44.19\% and
58.57\% of splice junctions from ESTs and ENSEMBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 15,399 (68.47\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score $\ge1.0$.
% @CTB
These matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

@@

\section*{Leftover Results text}

The program can merge all transcripts from multiple-kmers and predict
a structure of full-length isoforms and genes.
Total of 20,439 genes were obtained from global assembly and 21,963 genes
from local assembly.
However, the number of genes obtained from global + local assembly is only 20,922 genes.
The number of isoforms slightly increased after combining
transcripts from global and local assembly together (Table~\ref{genes_transcripts}).

Most transcripts include a large part of untranslated regions (UTRs), especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational methods due to low degree of conservation [?].
RNA-Seq-derived gene models are useful for studying variations within UTR regions, which may be involved in regulation of isoform expression\cite{} (Fig.~\ref{long_utr}).

\subsection*{An overestimated number of genes and transcripts}

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.  This seems to occur in a few genome
sequences such as chromosome E64\_random.
Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
gene models have only a few isoforms.

In addition, parameters used in building gene models such as minimum
exon/intron size determine how transcripts are merged together, which
can increases or decreases the number of isoforms.  In this study, we
adjust the parameters to enhance sensitivity of splice junctions
detection that may result in overestimation of the number of isoforms.

Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
especially $3'$ UTR in our datasets.
These regions are challenging to predict correctly solely from computational
methods due to low degree of conservation (cite).
RNA-Seq-derived gene models are useful for studying variations within UTR regions,
which may be involved in regulation of isoform expression (cite).
Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
$5'$UTRs and extended $3'$UTR respectively.

\subsubsection*{Cufflinks}

To evaluate an efficiency of our pipeline in detecting splice junctions,
we compared the number of splice junctions detected by our pipeline to
that from , ESTs and Ensembl gene models version 64.
Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
by our pipeline.
However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
by ESTs or mRNAs.
This suggests that the assembly detects some splice junctions that are
not detected by Cufflinks and they are genuine splice junctions.

We also aligned transcripts from Cufflinks and the pipeline to Ensembl
transcripts to compare genes and transcripts detected by both methods.
We selected only transcripts that match with more than 90\% identity and
cover more than 80\% of an Ensembl transcript.
Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
total 17,934 Ensembl genes.
6,798 of those genes are both detectd by our pipeline and Cufflinks.
In addition, we wanted to compare the number of isoforms detected by our
pipeline and Cufflinks within the same gene; however, a transcript can
match to multiple isoforms in the same gene, especially when it is not complete.
Therefore, we compare the number of splice junctions detected by both
methods within the same gene instead.
From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
Cufflinks detects 62,150 junctions.

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{RNA-Seq Assembly Discovers Many Splice Variants}
}
% Insert Author names, affiliations and corresponding author email.
\\
Likit Preeyanon$^{1}$, 
Jerry B. Dodgson$^{1}$
Hans Cheng$^{2}$, 
C. Titus Brown$^{3,1 \ast}$
\\
\bf{1} Microbiology and Molecular Genetics, Michigan State University, East Lansing, MI, USA.
\\
\bf{2} Avian Disease and Oncology Laboratory, East Lansing, MI, USA.
\\

\bf{3} Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, USA.
\\
$\ast$ E-mail: ctb@msu.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Splice variants play an important role in biological systems especially in the
immune system and the central nervous system. An expression profile of splice
variants has been shown to be a better signature of some diseases than an
overall gene expression profile~\cite{zhang2013isoform}. However, for most
organisms, available gene models such as Ensembl may not include all splice
variants expressed in RNA-Seq data. Reference-guided (e.g.
Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
models from RNA-Seq reads. However, we found that both types of method recovered
distinct set of splice variants.  In this article, we introduce a pipeline that
can combined transcripts from {\em de novo} assembly and other gene models to
constructs gene models that include more splice variants. We also described a
method called local assembly that we used to enhance the assembly of splice
variants to recover splice variants not found by Cufflinks and {\em de novo}
assembly.  Combination of Cufflinks gene models and transcripts from assembly
include more splice variants, increasing sensitivity of alternative splicing
study.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

% @CTB: aren't some splice variants detected only by Cufflinks and not by
% gimme, too?
% @LP: yes

% @CTB: for me to check, do Tophat and cufflinks only detect canonical splice?
% @LP: the mechanism used for reads >45 bp detects three canonical splice sites
% ab initio.

% @CTB: ask Olga for references to single-cell splice stuff.

@@

%Since sequences from RNA-Seq (reads) are very short (~50--250bp for Illumina
%reads), it is not feasible to assess expression isoforms and their structures
%without mapping reads to a reference annotation or reconstructing a full-length
%transcript from short reads.

@@

Using a similar approach, Pickrell {\em et al} identified more than
150,000 novel canonical splice junctions in lymphoblastoid cells.  The
study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests
tissue-specific expression of isoforms\cite{Pickrell:2010gt}. 
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
employing similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

Although these mapping-based methods have been useful in detecting
both splice junctions and isoforms, they rely heavily on a reference
genome.  Hence, it is not necessarily practical to apply these methods
to organisms lacking a high-quality reference genome.  This limitation
can be overcome by {\em de novo} assembly of short reads.

A number of {\em de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior than other {\em de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses {\em de novo} assembly
to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
apply a technique we call ``local assembly'' that enhances the sensitivity
of alternative isoform detection by Oases.  The results show that the
pipeline can detect more isoforms than Oases-M and can detect isoforms
not found by Cufflinks.  We also showed that transcripts reconstructed
from {\em de novo} assembly and mapping-based approach can be merged
to build more complete gene models.

@@

We used the Velvet\cite{Zerbino:2008vu} and Oases\cite{Schulz:2012je}
assemblers to construct transcript fragments from four entire Illumina
GAII mRNAseq data sets sequenced from chicken spleen (see Methods and
Materials).  In the assembly, we used multiple distinct k-mer values
for Velvet to sensitively recover as many different isoforms as possible
\cite{Schulz:2012je}.  We chose k-mers between 21 and 31,
and recovered between 90,700 and 104,000 unique sequences from
each data set (see Table~\ref{total_sequences}, and Materials and
Methods).  These unique sequences represented an unknown number of
true genes, due to fragmentation from low coverage and incomplete
assembly.

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly (Table~\ref{unique_sequences},

% @CTB why are the global ones not included in the final gene models?
% @LP they are included in the final models if they align to the genome.

and Table~\ref{unique_sequences_matched_mouse}); this is probably due to the
incomplete nature of the current chicken genome assembly, which is lacking
approximately 5\% of its true gene content (@cite).  Interestingly, over a
hundred regions were present {\em only} in the local assemblies,
suggesting that the local assemblies might be recovering additional exons.
Significant numbers of unique regions from both global and local assemblies
showed homology to the mouse genome, indicating that at least some of these
unique sequences represented real sequence content.
Figure~\ref{global_vs_local} shows an example of different isoforms detected by
the two assembly methods.

% @CTB how many reads are discarded in the process of mapping?

% @CTB what does local assembly actually do to the DBG??
% @LP I think the difference is that ambiguous reads, especially spliced reads
% are used for assembly more than once in local assembly because Tophat reports
% multiple alignments. So, reads mapped to multiple chromosomes get assembled
% independently. Assembling all mapped reads at once misses some splice
% junctions when compared to assembling each chromosome separately.

% @CTB what happens if you ask how many unique regions there are not
%   represented in the final gene models?

@@

The above approaches recovered transcript fragments, but not entire genes
(figure in suppl?)
To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.

% @CTB what does this mean? ``not fully evaluated''?
% @LP Oases paper only evaluate completeness of transcripts by their alignments
% to the ref sequences. I believe the cutoff they used for a good alignment was
% 80 percent match. I have to reread it again though, but I'm quite positive.

We merged transcripts from our global assembly, above, with Oases-M using a
k-mer size of 27, and compared them with the unmerged transcripts.  We then
cross-validated using publicly available ESTs, which were not used in our
assembly.  The results show that Oases-M and the unmerged assembly share about
104,413 (94\%) of the predicted splice junctions, with 6\% disjoint.  Of these
6\%, approximately 420 (6.1\%) of the Oases-M-specific splice junctions are
independently supported by ESTs, while 1,608 (19.4\%) of the unmerged splice
junctions are supported by ESTs (Table~\ref{Oases-M}).  This suggests that Oases-M
probably discards a number of real splice variants, although the unmerged
assembly is also missing some found by Oases-M.

% @CTB why would Oases-M create ``good'' splice variants from the unmerged
% assembly?
% @LP My guess is OasesM can extend transcripts from all k-mers and longer
% transcripts may be aligned better to the genome.

% @CTB does the math for table three work out here?
% @LP yes, percent match is calculated from unique region of each type of
% assembly.

% @CTB is Oases-M meant to work with non-oases-assembled transcripts? is it a
% general merging tool?
% @LP I think it's kind of a general merging tool.

@@

We used an exon graph approach to construct gene models from
alignments of transcripts against the genome.  Our approach,
implemented in a software package called Gimme, merges transcripts and
gene models based on overlapping exons using an exon-graph approach
(see Materials and Methods; Figure~\ref{algorithm}).  We used Gimme to obtain
21,492 gene models containing 31,418 isoforms from our global
assembly; 24,928 gene models containing 29,776 from our local
assembly; and 22,587 gene models containing 34,800 isoforms from the
merged global and local assemblies (Table~\ref{genes_transcripts}).

@@

Our pipeline predicts many gene models and isoforms after assembly.  We
validated these gene models in several different ways.

@@

We used Bowtie to map single-end reads from the source datasets to the
transcripts.  More than $75-80$\% of the original reads could be aligned to the
transcripts, demonstrating that we did not lose a significant amount of
information during the merge process compared to the number of reads mapped to
Ensembl gene models (Figure~\ref{mapped-reads}).

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse transcription (@cite).

% @CTB do we have something we can cite for typical mapping?  Maybe Trinity paper?
% How about a citation that suggests that there's a lot of crap in mRNAseq?
% @LP I'll search for it.

Thus the merged gene models produced by Gimme represent the significant majority
of the assemblable data.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in random chromosomes,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).

% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.

More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.

% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the reason I use 4 as a cutoff because there are four samples and if at
% least one read from each sample map to the junction, it could be real.
% Apparently, it is not always the case, but that is my simple logic.

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank~\ref{chick_est_venn}.  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

% @CTB supported by ESTs OR mRNAs, or ESTs AND mRNAs?  Surely the former?
% @LP supported by ESTs or mRNAs. I unioned splice junctions from ESTs and
% mRNAs.

@@

Cufflinks finds 92,077 splice junctions, and Gimme finds 105,461 splice
junctions.  80,964 of them are in common (Figure~\ref{chick_venn}). Both
Cufflinks and Gimme find approximately 40-50\% of the genes and 50-53\% of the
splice junctions present in the ENSEMBL gene models for chicken.  The ENSEMBL
pipeline does not, however, include a large number of splice junctions from ESTs
(97,740) or mRNAs (13,987).  Cufflinks and Gimme each recover about 18\% of
these, with more than 2/3 of these recovered by both Cufflinks and Gimme.  This
indicates that both Gimme and Cufflinks are equally adept at recovering novel
splice junctions.

% @CTB note this suggests that Cufflinks improved recovery for ENSEMBL
% gene models is false...
% @LP it does recover some that not recovered by assembly.
% @CTB what fraction of Cufflinks splice junctions are high coverage (4+)?
% @LP a bit higher than our gene models, around 99\%, 179 junctions with no
% reads and 1010 junctions <4.

@@

When we apply Gimme and Cufflinks to a publicly available mouse mRNAseq data
set, Gimme and Cufflinks recover approximately the same number of splice
junctions already known from ENSEMBL (Figure ~\ref{mus_venn}).  However, Gimme
recovers a substantial number of additional splice variants beyond Cufflinks and
ENSEMBL both.

% @CTB is the mouse data set from the trinity paper?
% @LP yes

% @CTB are most of these splice junctions > 3 coverage?
% @LP I'll figure that out later on.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Figure~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan successfully
translated 28,772 of 34,800 (82.6\%) of our isoforms to protein sequences with
50 or more amino acids.  We then searched for homologous sequences in mouse
ENSEMBL, and found that 22,991 (79.9\%) of our isoforms from 12,945 distinct
genes match mouse proteins at a bit score $\ge1.0$.
These matches have a bit-score/length ratio greater than 1, which indicates a
good agreement between chicken and mouse proteins.

@@

% \section*{Leftover Results text}
% 
% The program can merge all transcripts from multiple-kmers and predict a
% structure of full-length isoforms and genes.  Total of 20,439 genes were
% obtained from global assembly and 21,963 genes from local assembly.  However,
% the number of genes obtained from global + local assembly is only 20,922 genes.
% The number of isoforms slightly increased after combining transcripts from
% global and local assembly together (Table~\ref{genes_transcripts}). 
% 
% Most transcripts include a large part of untranslated regions (UTRs), especially
% $3'$ UTR in our datasets.  These regions are challenging to predict correctly
% solely from computational methods due to low degree of conservation [?].
% RNA-Seq-derived gene models are useful for studying variations within UTR
% regions, which may be involved in regulation of isoform expression\cite{}
% (Fig.~\ref{long_utr}).
% 
% \subsection*{An overestimated number of genes and transcripts}
% 
% The number of genes from our pipeline might be overestimated due to
% fragmented transcripts.  In our datasets, read coverage is lower at
% the $5'$ end due to the method used to convert mRNA to cDNA in library
% preparation.  A bias of read coverage toward $3'$ end and low
% expression level result in fragmented transcripts as shown in
% Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
% junctions due to duplication in genome sequence can lead to a large
% number of dubious isoforms.  This seems to occur in a few genome
% sequences such as chromosome E64\_random.
% Figure~\ref{isoforms_boxplot} shows that a majority of genes from our
% gene models have only a few isoforms.
% 
% In addition, parameters used in building gene models such as minimum
% exon/intron size determine how transcripts are merged together, which
% can increases or decreases the number of isoforms.  In this study, we
% adjust the parameters to enhance sensitivity of splice junctions
% detection that may result in overestimation of the number of isoforms.
% 
% Most transcripts from RNA-Seq assembly include untranslated regions (UTRs),
% especially $3'$ UTR in our datasets.
% These regions are challenging to predict correctly solely from computational
% methods due to low degree of conservation (cite).
% RNA-Seq-derived gene models are useful for studying variations within UTR regions,
% which may be involved in regulation of isoform expression (cite).
% Figure~\ref{alter_5utr} and figure~\ref{long_utr} shows an example of alternative
% $5'$UTRs and extended $3'$UTR respectively.
% 
% \subsubsection*{Cufflinks}
% 
% To evaluate an efficiency of our pipeline in detecting splice junctions,
% we compared the number of splice junctions detected by our pipeline to
% that from , ESTs and Ensembl gene models version 64.
% Cufflinks detected 88,289 (87.93\%) splice junctions that are also detected
% by our pipeline.
% However, 42.43\% (5,143) of junctions not detected by Cufflinks are supported
% by ESTs or mRNAs.
% This suggests that the assembly detects some splice junctions that are
% not detected by Cufflinks and they are genuine splice junctions.
% 
% We also aligned transcripts from Cufflinks and the pipeline to Ensembl
% transcripts to compare genes and transcripts detected by both methods.
% We selected only transcripts that match with more than 90\% identity and
% cover more than 80\% of an Ensembl transcript.
% Our pipeline detects 7,298 genes and Cufflinks detects 8,341 genes of
% total 17,934 Ensembl genes.
% 6,798 of those genes are both detectd by our pipeline and Cufflinks.
% In addition, we wanted to compare the number of isoforms detected by our
% pipeline and Cufflinks within the same gene; however, a transcript can
% match to multiple isoforms in the same gene, especially when it is not complete.
% Therefore, we compare the number of splice junctions detected by both
% methods within the same gene instead.
% From 6,798 genes, our pipeline detects 56,332 splice junctions, whereas
% Cufflinks detects 62,150 junctions.
% 
% However, because Ensembl gene models do not include all isoforms or
% splice junctions from ESTs, we investigated splice junctions that are
% not in Ensembl but are supported by ESTs in those 6,798 genes and
